---
title: "DS210 Project Proposal"
author: "Laina Cleaton"
date: "2025-10-28"
output: 
  html_document:
    theme: cosmo 
---

**Section 1 - Introduction:**

Customer retention is a critical concern in the banking industry, where competition and customer expectations are both high. Understanding why customers choose to leave a particular bank, or "churn", can provide valuable insights for improving bank services and maintaining customer satisfaction. 

During my internship at Bank of America, I supported the Call Insights function in the Retail Transformation team, focusing on data analytics to enhance the client experience. I enjoyed identifying trends in customer call and complaint data to improve bank service efficiency. This project allows me to build on those skills by applying similar analytical techniques to a different dataset within the banking industry. 

*Research Question: What factors are most strongly associated with whether or not a customer closes their account with a bank?*

The purpose of this project is to explore the factors that influence customer churn, with the goal of identifying patterns and relationships between customer characteristics and their decision to close their account. The dataset used in this study, titled "*Bank Customer Churn*," is publicly available from [Maven Analytics' Data Playground](https://mavenanalytics.io/data-playground/bank-customer-churn?page=7&pageSize=5), which sourced the data from [Kaggle](https://www.kaggle.com/datasets/shubhammeshram579/bank-customer-churn-prediction/data). While the specific bank is not identified, the dataset represents customer information from an unnamed European bank and contains 10,000 observations. 

Each of the 10,000 observations represents a single customer and includes 13 variables capturing demographic, behavioral, and bank account information. These variables include *CustomerId* (unique identifier), *Surname* (customer's last name), *CreditScore* (customer's credit score), *Geography* (the customer's country of residence: France, Germany, or Spain), *Gender* (customer's sex: Male or Female), *Age* (customer's age in years), *Tenure* (number of years as a customer with the bank), *Balance* and *EstimatedSalary* (customer's account balance and estimated salary, both in euros, €), *HasCrCard* (whether or not the customer has a credit card with the bank), *NumOfProducts* (the number of bank products the customer owns, such as a debit card), and *IsActiveMember* (whether the customer actively engages with the bank). The outcome variable, *Exited*, indicates whether or not the customer has closed their account with the bank. 

The data on each customer was collected from the bank's internal records, which means that each row corresponds to one customer's status at the time the data was collected. In this analysis, the variable *Exited* is the dependent variable representing customer churn, while the remaining variables serve as potential predictors of the decision to churn. 

By examining this dataset, the project aims to uncover trends and key factors that contribute to customer churn. These insights can inform strategies to improve customer retention and guide future efforts in customer relationship management. 


**Section 2 - Exploratory Data Analysis:**

The purpose of this exploratory data analysis is to understand the structure and characteristics of the *Bank Customer Churn* dataset, identify any potential data quality issues, and explore initial relationships between variables that may help explain customer churn. The ultimate goal is to identify patterns associated with the response variable, *Exited*. 

***Step 1: Import and Inspect the Dataset***

```{r setup_tidyverse, eval=FALSE}
library(tidyverse)
```

```{r read_file, eval = FALSE}
bank_churn_data <- read_csv("Bank_Churn.csv")
View(bank_churn_data)
str(bank_churn_data)
head(bank_churn_data)
```

The dataset contains 10,000 observations and 13 variables.

***Step 2: Data Cleaning and Preparation***

The variables *CustomerId* and *Surname* are identifiers and do not contribute to the analysis, so they are removed.

```{r clean_dataset, eval = FALSE}
bank_churn_clean <- bank_churn_data %>%
  select(-CustomerId, -Surname)
View(bank_churn_clean)
```

The binary variables *HasCrCard*, *IsActiveMember*, and *Exited* are converted from numeric (1 or 0) to categorical ('Yes' or 'No') values for clarity in analysis and visualization.

```{r recode_binary, eval = FALSE}
bank_churn_clean <- bank_churn_clean %>%
  mutate(
    HasCrCard = ifelse(HasCrCard == "1", "Yes", "No"),
    IsActiveMember = ifelse(IsActiveMember == "1", "Yes", "No"),
    Exited = ifelse(Exited == "1", "Yes", "No")
  )
View(bank_churn_clean)
```


***Step 3: Checking for Missing Data***

It is important to verify that all variables contain complete information, since missing values can distort statistical results.

```{r na_check, eval = FALSE}
colSums(is.na(bank_churn_clean))
```
All variables contain complete information, so no further cleaning is required.

***Step 4: Individual Variable Summary Statistics and Visualizations***


**Credit Score**

```{r creditscore_summary_visualization, eval = FALSE}
summary(bank_churn_clean$CreditScore)
sd(bank_churn_clean$CreditScore)

ggplot(bank_churn_clean, aes(x = CreditScore)) +
  geom_histogram(binwidth = 25, fill = "steelblue", color = "white") +
  labs(title = "Distribution of Credit Score", subtitle = "For Customers of a European Bank", x = "Credit Score", y = "Count") +
  theme_minimal()
```
The shape of the Credit Score distribution is fairly normal without any skew, so we can use the mean for the measure of center and standard deviation and range for the measures of spread. The distribution has a mean Credit Score of 650.5, a standard deviation of 96.6533, and a range of 500. 

**Age**

```{r age_summary_visualization, eval = FALSE}
summary(bank_churn_clean$Age)
sd(bank_churn_clean$Age)

ggplot(bank_churn_clean, aes(x = Age)) +
  geom_histogram(binwidth = 5, fill = "#5F9EA0", color = "white") +
  labs(title = "Distribution of Age", subtitle = "For Customers of a European Bank", x = "Age", y = "Count") +
  theme_minimal()
```
The shape of the Customer Age distribution is slightly right skewed, so we will use the median for the measure of center and standard deviation and IQR for the measures of spread. The distribution has a median Age of 37 years, a standard deviation of 10.48781 years, and an IQR of 12 years.

**Tenure**

```{r tenure_summary_visualization, eval = FALSE}
summary(bank_churn_clean$Tenure)
sd(bank_churn_clean$Tenure)

ggplot(bank_churn_clean, aes(x = Tenure)) +
  geom_histogram(binwidth = 1, fill = "#B0E0E6", color = "white") +
  labs(title = "Distribution of Tenure", subtitle = "For Customers of a European Bank", x = "Tenure", y = "Count") +
  theme_minimal()
```
The shape of the Tenure distribution is fairly uniform, so we can use the median or the mean for the measure of center and standard deviation and range for the measures of spread. The distribution has a median Tenure of 5 years, a mean Tenure of 5.013 years, a standard deviation of 2.892174 years, and an IQR of 4 years.

**Number of Products**
``` {r numproducts_summary_visualization, eval = FALSE}
table(bank_churn_clean$NumOfProducts)
sd(bank_churn_clean$NumOfProducts)

ggplot(bank_churn_clean, aes(x = NumOfProducts)) +
  geom_histogram(binwidth = 1, fill = "#4682B4", color = "white") +
  labs(title = "Distribution of Number of Products", subtitle = "For Customers of a European Bank", x = "Number of Products", y = "Count") +
  theme_minimal()
```
5,084 customers own 1 product, 4,590 customers own 2 products, 266 customers own 3 products, and 60 customers own 4 products from the European bank. The majority of customers own 1 product. 

**Account Balance**

```{r balance_summary_visualization, eval = FALSE}
library(scales)

summary(bank_churn_clean$Balance)
sd(bank_churn_clean$Balance)

ggplot(bank_churn_clean, aes(x = Balance)) +
  geom_histogram(binwidth = 10000, fill = "#008080", color = "white") +
  labs(title = "Distribution of Account Balance (in euros)", subtitle = "For Customers of a European Bank", x = "Balance (€)", y = "Count") +
  scale_x_continuous(labels = comma) +
  theme_minimal()
```
The distribution is fairly normal, except for the very large and outlier of customers having a $0 balance in their bank accounts. I think that this reflects those who have closed their account, and thus do not have any money still in their closed accounts. Due to this large left skew, we will use the median for the measure of center and standard deviation and IQR for the measures of spread. The distribution has a median Account Balance of 97,199 euros, a standard deviation of 62,397.41 euros, and an IQR of 127,644 euros.

**Estimated Salary**

```{r estimatedsalary_summary_visualization, eval = FALSE}
library(scales)

summary(bank_churn_clean$EstimatedSalary)
sd(bank_churn_clean$EstimatedSalary)

ggplot(bank_churn_clean, aes(x = EstimatedSalary)) +
  geom_histogram(binwidth = 7500, fill = "lightblue", color = "white") +
  labs(title = "Distribution of Estimated Salary (in euros)", subtitle = "For Customers of a European Bank", x = "Estimated Salary (€)", y = "Count") +
  scale_x_continuous(labels = comma) +
  theme_minimal()
```
The shape of the Estimated Salary distribution is fairly uniform, so we can use the median or the mean for the measure of center and standard deviation and range for the measures of spread. The distribution has a median Estimated Salary of 100,193.91 euros, a mean Estimated Salary of 100,090.24 euros, a standard deviation of 57,510.49 euros, and an IQR of 98,386.14 euros.

**Country of Residence**

```{r geography_summary_visualization, eval = FALSE}
table(bank_churn_clean$Geography)

ggplot(bank_churn_clean, aes(x = Geography)) +
  geom_bar(fill = "steelblue") +
  labs(title = "Distribution of Country of Residence", subtitle = "For Customers of a European Bank", x = "Country", y = "Count") +
  theme_minimal()
```

5,014 customers reside in France, 2,509 customers reside in Germany, and 2,477 customers reside in Spain. The majority of customers reside in France. 

**Gender**

```{r gender_summary_visualization, eval = FALSE}
table(bank_churn_clean$Gender)

ggplot(bank_churn_clean, aes(x = Gender)) +
  geom_bar(fill = "#5F9EA0") +
  labs(title = "Number of Customers by Gender", subtitle = "For Customers of a European Bank", x = "Gender", y = "Count") +
  theme_minimal()
```
4,543 customers are female and 5,457 customers are male. The majority of the customers of the European bank are male. 

**Credit Card Ownership**
```{r creditcard_summary_visualization, eval = FALSE}

table(bank_churn_clean$HasCrCard)

ggplot(bank_churn_clean, aes(x = HasCrCard)) +
  geom_bar(fill = "#B0E0E6") +
  labs(title = "Credit Card Ownership", subtitle = "For Customers of a European Bank", x = "Owns a Credit Card", y = "Count") +
  theme_minimal()
```
2,945 customers do not own a credit card, and 7,055 custoemrs do own a credit card. The majority of customers own a credit card. 

**Active Bank Membership Status**
```{r activemember_summary_visualization, eval = FALSE}

table(bank_churn_clean$IsActiveMember)

ggplot(bank_churn_clean, aes(x = IsActiveMember)) +
  geom_bar(fill = "#4682B4") +
  labs(title = "Active Bank Membership Status", subtitle = "For Customers of a European Bank", x = "Is Active Member of Bank", y = "Count") +
  theme_minimal()
```
4,849 of the customers are not active members of the bank, whereas 5,151 customers are active. This makes me wonder if about half of the customers have left the bank.

**Customer Churn**
```{r exited_summary_visualization, eval = FALSE}

table(bank_churn_clean$Exited)

ggplot(bank_churn_clean, aes(x = Exited)) +
  geom_bar(fill = "#008080") +
  labs(title = "Customer Churn", subtitle = "For Customers of a European Bank", x = "Exited", y = "Count") +
  theme_minimal()
```
Surprisingly, despite almost half of the customers not being active in the bank, only 2,037 customers have churned. 7,963 customers have not left the bank. 

***Step 5: Two Variable Comparisons and Visualizations***

**Churn Status by Credit Score**
```{r creditscore_churn_visualization, eval = FALSE}
ggplot(bank_churn_clean, aes(x = as.numeric(CreditScore), y = factor(Exited))) +
  geom_violin(fill = "#B0E0E6", alpha = 0.6) +
  geom_boxplot(fill = "#5F9EA0") +
  labs(title = "Churn Status by Credit Score", subtitle = "For Customers of a European Bank", x = "Credit Score", y = "Exited") +
  theme_minimal()
```
Credit Score does not appear to have a significant impact on whether a customer left the bank or not. The boxplots have similar medians and spreads. 

**Churn Status by Age**
```{r age_churn_visualization, eval = FALSE}
ggplot(bank_churn_clean, aes(x = as.numeric(Age), y = factor(Exited))) +
  geom_violin(fill = "lightblue", alpha = 0.6) +
  geom_boxplot(fill = "steelblue") +
  labs(title = "Churn Status by Age", subtitle = "For Customers of a European Bank", x = "Age", y = "Exited") +
  theme_minimal()
```
Of those who left the bank, the median age appears to be about 43 years old. The distribution of the age of those who left the bank appears to be slightly right skewed. I want to explore the characteristics of those of an older age, and see if I can uncover any trends. 

**Churn Status by Tenure**
```{r tenure_churn_visualization, eval = FALSE}
ggplot(bank_churn_clean, aes(x = as.numeric(Tenure), y = factor(Exited))) +
  geom_boxplot(fill = "#008080") +
  labs(title = "Churn Status by Tenure", subtitle = "For Customers of a European Bank", x = "Tenure", y = "Exited") +
  theme_minimal()
```
Those who exited the bank have a larger spread of tenure than those who did not. It is quite interesting, but does not help to uncover why people left the bank. 

**Churn Status by Number of Products**
```{r numproducts_churn_visualization, eval = FALSE}
ggplot(bank_churn_clean, aes(x = factor(NumOfProducts), fill = Exited)) +
         geom_bar(position = "fill") +
  labs(title = "Churn Status by Number of Products", subtitle = "For Customers of a European Bank", x = "Number of Products", y = "Proportion of Customers") +
  scale_fill_manual(values = c("lightblue", "#5F9EA0"), name = "Exited") +
  theme_minimal()
```
Although only 60 people had 4 products in the bank, it is fascinating that 100% of those 60 customers left the bank. Over 75% of those that owned 3 products also left the bank. I want to explore data with just those two variables.

**Churn Status by Account Balance**
```{r balance_churn_visualization, eval = FALSE}
ggplot(bank_churn_clean, aes(x = as.numeric(Balance), y = factor(Exited))) +
  geom_violin(fill = "lightblue", alpha = 0.6) +
  geom_boxplot(fill = "#4682B4") +
  labs(title = "Churn Status by Balance (in euros)", subtitle = "For Customers of a European Bank", x = "Balance (€)", y = "Exited") +
  theme_minimal()
```
A higher Account Balance median occurs for those who left the bank than for those who did not. I want to see why this is the case, and if it relates to the Age or Number of Products results. 

**Churn Status by Estimated Salary**
```{r estsalary_churn_visualization, eval = FALSE}
ggplot(bank_churn_clean, aes(x = as.numeric(EstimatedSalary), y = factor(Exited))) +
  geom_violin(fill = "lightblue", alpha = 0.6) +
  geom_boxplot(fill = "#4682B4") +
  labs(title = "Churn Status by Estimated Salary (in euros)", subtitle = "For Customers of a European Bank", x = "Estimated Salary (€)", y = "Exited") +
  theme_minimal()
```
Estimated Salary does not appear to have a significant impact on whether a customer left the bank or not. The boxplots have similar medians and spreads. 

**Churn Status by Country of Residence**
```{r geography_churn_visualization, eval = FALSE}

ggplot(bank_churn_clean, aes(x = Geography, fill = Exited)) +
  geom_bar(position = "fill") +
  labs(title = "Churn Status by Country of Residence", subtitle = "For Customers of a European Bank", x = "Country", y = "Proportion") +
  scale_fill_manual(values = c("lightblue", "#5F9EA0"), name = "Exited")
  theme_minimal()
```
Although the majority of customers reside in France, the highest proportion of those who left the bank, at over 0.25, resided in Germany. 

**Churn Status by Gender**
```{r gender_churn_visualization, eval = FALSE}

ggplot(bank_churn_clean, aes(x = Gender, fill = Exited)) +
  geom_bar(position = "fill") +
  labs(title = "Churn Status by Gender", subtitle = "For Customers of a European Bank", x = "Gender", y = "Proportion") +
  scale_fill_manual(values = c("lightblue", "#5F9EA0"), name = "Exited")
  theme_minimal()
```
More female customers have left the bank than male customers, though it is by about 10%. 

**Churn Status by Credit Card Ownership**

```{r creditcard_churn_visualization, eval = FALSE}

ggplot(bank_churn_clean, aes(x = HasCrCard, fill = Exited)) +
  geom_bar(position = "fill") +
  labs(title = "Churn Status by Credit Card Ownership", subtitle = "For Customers of a European Bank", x = "Owns a Credit Card", y = "Proportion") +
  scale_fill_manual(values = c("lightblue", "#5F9EA0"), name = "Exited")
  theme_minimal()
```
Whether a customer owned a credit card or not does not appear to have a significant impact on whether a customer left the bank or not. The proportions are almost equivalent. 

**Churn Status by Active Bank Membership Status**
```{r activemember_churn_visualization, eval = FALSE}

ggplot(bank_churn_clean, aes(x = IsActiveMember, fill = Exited)) +
  geom_bar(position = "fill") +
  labs(title = "Churn Status by Active Bank Membership Status", subtitle = "For Customers of a European Bank", x = "Is Active Member of Bank", y = "Proportion") +
  scale_fill_manual(values = c("lightblue", "#5F9EA0"), name = "Exited")
  theme_minimal()
```
I would have expected a higher proportion of those who exited the bank to no longer be an active member of the bank. However, it does make sense that a higher proportion of those who churned would no longer be active. 

***Step 5: Multivariable Visualizations and Analysis***

**Age vs. Balance by Exited**
```{r age_balance_churn, eval = FALSE}
bank_churn_clean <- bank_churn_clean %>%
  mutate(AgeGroup = case_when(
    Age < 30 ~ "<30",
    Age >= 30 & Age < 40 ~ "30–39",
    Age >= 40 & Age < 50 ~ "40–49",
    Age >= 50 & Age < 60 ~ "50–59",
    TRUE ~ "60+"
  ))

ggplot(bank_churn_clean, aes(x = Balance, y = AgeGroup, fill = Exited)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  labs(
    title = "Account Balance by Age Group and Churn Status",
    subtitle = "For Customers of a European Bank",
    x = "Balance (€)",
    y = "Age Group",
    fill = "Exited"
  ) +
  scale_x_continuous(labels = scales::comma) +
  scale_fill_manual(values = c("No" = "lightblue", "Yes" = "#5F9EA0")) +
  theme_minimal()
```
I wanted to see the relationship between age groups and the amount of money they have in their bank accounts, and whether that influenced bank churns. It is interesting that there is a similar theme across 

**Customer Churn by Number of Products and Country of Residence**

```{r germany_numproducts_churn, eval = FALSE}

ggplot(bank_churn_clean,aes(x = factor(NumOfProducts), fill = Exited)) +
  geom_bar(position = "fill") +
  scale_fill_manual(values = c("No" = "lightblue", "Yes" = "#5F9EA0")) +
  facet_wrap(~ Geography) +
  labs(
    title = "Customer Churn by Number of Products and Country of Residence",
    subtitle = "For Customers of a European Bank",
    x = "Number of Products",
    y = "Proportion of Customers",
    fill = "Exited"
  ) +
  theme_minimal()
```
As I suspected, a higher proportion of those residing in Germany exited the bank when they had 1, 2, and 3 products than in any of the other two countries. All countries obviously had 100% of customers with 4 products leave the bank. 

**Average Balance by Age for Customers with 3 or 4 Products**

```{r balance_age_churn, eval = FALSE}
bank_churn_clean %>%
  filter(Exited == "Yes") %>%
  mutate(AgeGroup = cut(Age, breaks = seq(20, 80, by = 5))) %>%  # 5-year bins
  group_by(AgeGroup, NumOfProducts) %>%
  summarise(avg_balance = mean(Balance, na.rm = TRUE), .groups = "drop") %>%
  ggplot(aes(x = AgeGroup, y = avg_balance, color = factor(NumOfProducts), group = NumOfProducts)) +
  geom_line(linewidth = 1.2) +
  geom_point() +
  scale_y_continuous(labels = comma) +
 scale_color_manual(
    values = c("#5F9EA0", "#4682B4", "lightblue", "#008080"),
    name = "Number of Products"
  ) +
  labs(
    title = "Average Balance by Age Group and Number of Products",
    subtitle = "For Churned Customers of a European Bank",
    x = "AgeGroup",
    y = "Average Balance (€)"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5))
```
An interesting trend that we can see is that the most of customers who reached 70 years old did not leave the bank, except for those who owned 3 products. I wish I knew which products resulted in people leaving, as perhaps the technology of the 3rd product was too advanced for many customers, especially those of an older age. Perhaps an online banking product with an AI component. But of what we know, we can see the largest spike in people leaving the bank when they own three products and are above the age of 65-70. 
