---
title: "DS210 Project Proposal"
author: "Laina Cleaton"
date: "2025-10-28"
output: 
  html_document:
    theme: cosmo 
---

**Section 1 - Introduction:**

Customer retention is a critical concern in the banking industry, where competition and customer expectations are both high. Understanding why customers choose to leave a particular bank, or "churn", can provide valuable insights for improving bank services and maintaining customer satisfaction. 

During my internship at Bank of America, I supported the Call Insights function in the Retail Transformation team, focusing on data analytics to enhance the client experience. I enjoyed identifying trends in customer call and complaint data to improve bank service efficiency. This project allows me to build on those skills by applying similar analytical techniques to a different dataset within the banking industry. 

*Research Question: What factors are most strongly associated with whether or not a customer closes their account with a bank?*

The purpose of this project is to explore the factors that influence customer churn, with the goal of identifying patterns and relationships between customer characteristics and their decision to close their account. The dataset used in this study, titled "*Bank Customer Churn*," is publicly available from [Maven Analytics' Data Playground](https://mavenanalytics.io/data-playground/bank-customer-churn?page=7&pageSize=5), which sourced the data from [Kaggle](https://www.kaggle.com/datasets/shubhammeshram579/bank-customer-churn-prediction/data). While the specific bank is not identified, the dataset represents customer information from an unnamed European bank and contains 10,000 observations. 

Each of the 10,000 observations represents a single customer and includes 13 variables capturing demographic, behavioral, and bank account information. These variables include *CustomerId* (unique identifier), *Surname* (customer's last name), *CreditScore* (customer's credit score), *Geography* (the customer's country of residence: France, Germany, or Spain), *Gender* (customer's sex: Male or Female), *Age* (customer's age in years), *Tenure* (number of years as a customer with the bank), *Balance* and *EstimatedSalary* (customer's account balance and estimated salary, both in euros, €), *HasCrCard* (whether or not the customer has a credit card with the bank), *NumOfProducts* (the number of bank products the customer owns, such as a debit card), and *IsActiveMember* (whether the customer actively engages with the bank). The outcome variable, *Exited*, indicates whether or not the customer has closed their account with the bank. 

The data on each customer was collected from the bank's internal records, which means that each row corresponds to one customer's status at the time the data was collected. In this analysis, the variable *Exited* is the dependent variable representing customer churn, while the remaining variables serve as potential predictors of the decision to churn. 

By examining this dataset, the project aims to uncover trends and key factors that contribute to customer churn. These insights can inform strategies to improve customer retention and guide future efforts in customer relationship management. 


**Section 2 - Exploratory Data Analysis:**

The purpose of this exploratory data analysis is to understand the structure and characteristics of the *Bank Customer Churn* dataset, identify any potential data quality issues, and explore initial relationships between variables that may help explain customer churn. The ultimate goal is to identify patterns associated with the response variable, *Exited*. 

***Step 1: Import and Inspect the Dataset***

```{r setup_tidyverse, eval=FALSE}
library(tidyverse)
```

```{r read_file, eval = FALSE}
bank_churn_data <- read_csv("Bank_Churn.csv")
View(bank_churn_data)
str(bank_churn_data)
head(bank_churn_data)
```

The dataset contains 10,000 observations and 13 variables.

***Step 2: Data Cleaning and Preparation***

The variables *CustomerId* and *Surname* are identifiers and do not contribute to the analysis, so they are removed.

```{r clean_dataset, eval = FALSE}
bank_churn_clean <- bank_churn_data %>%
  select(-CustomerId, -Surname)
View(bank_churn_clean)
```

The binary variables *HasCrCard*, *IsActiveMember*, and *Exited* are converted from numeric (1 or 0) to categorical ('Yes' or 'No') values for clarity in analysis and visualization.

```{r recode_binary, eval = FALSE}
bank_churn_clean <- bank_churn_clean %>%
  mutate(
    HasCrCard = ifelse(HasCrCard == "1", "Yes", "No"),
    IsActiveMember = ifelse(IsActiveMember == "1", "Yes", "No"),
    Exited = ifelse(Exited == "1", "Yes", "No")
  )
View(bank_churn_clean)
```


***Step 3: Checking for Missing Data***

It is important to verify that all variables contain complete information, since missing values can distort statistical results.

```{r na_check, eval = FALSE}
colSums(is.na(bank_churn_clean))
```
All variables contain complete information, so no further cleaning is required.

***Step 4: Individual Variable Summary Statistics and Visualizations***


**Credit Score**

```{r creditscore_summary_visualization, eval = FALSE}
summary(bank_churn_clean$CreditScore)
sd(bank_churn_clean$CreditScore)

ggplot(bank_churn_clean, aes(x = CreditScore)) +
  geom_histogram(binwidth = 25, fill = "steelblue", color = "white") +
  labs(title = "Distribution of Credit Score", subtitle = "For Customers of a European Bank", x = "Credit Score", y = "Count") +
  theme_minimal()
```
(analysis here)


**Age**

```{r age_summary_visualization, eval = FALSE}
summary(bank_churn_clean$Age)
sd(bank_churn_clean$Age)

ggplot(bank_churn_clean, aes(x = Age)) +
  geom_histogram(binwidth = 5, fill = "#5F9EA0", color = "white") +
  labs(title = "Distribution of Age", subtitle = "For Customers of a European Bank", x = "Age", y = "Count") +
  theme_minimal()
```
(analysis here)

**Tenure**

```{r tenure_summary_visualization, eval = FALSE}
summary(bank_churn_clean$Tenure)
sd(bank_churn_clean$Tenure)

ggplot(bank_churn_clean, aes(x = Tenure)) +
  geom_histogram(binwidth = 1, fill = "#B0E0E6", color = "white") +
  labs(title = "Distribution of Tenure", subtitle = "For Customers of a European Bank", x = "Tenure", y = "Count") +
  theme_minimal()
```
(analysis here)

**Number of Products**
``` {r numproducts_summary_visualization, eval = FALSE}
summary(bank_churn_clean$NumOfProducts)
sd(bank_churn_clean$NumOfProducts)

ggplot(bank_churn_clean, aes(x = NumOfProducts)) +
  geom_histogram(binwidth = 1, fill = "#4682B4", color = "white") +
  labs(title = "Distribution of Number of Products", subtitle = "For Customers of a European Bank", x = "Number of Products", y = "Count") +
  theme_minimal()
```

**Account Balance**

```{r balance_summary_visualization, eval = FALSE}
library(scales)

summary(bank_churn_clean$Balance)
sd(bank_churn_clean$Balance)

ggplot(bank_churn_clean, aes(x = Balance)) +
  geom_histogram(binwidth = 10000, fill = "#008080", color = "white") +
  labs(title = "Distribution of Account Balance (in euros)", subtitle = "For Customers of a European Bank", x = "Balance (€)", y = "Count") +
  scale_x_continuous(labels = comma) +
  theme_minimal()
```
(analysis)

**Estimated Salary**

```{r estimatedsalary_summary_visualization, eval = FALSE}
library(scales)

summary(bank_churn_clean$EstimatedSalary)
sd(bank_churn_clean$EstimatedSalary)

ggplot(bank_churn_clean, aes(x = EstimatedSalary)) +
  geom_histogram(binwidth = 7500, fill = "lightblue", color = "white") +
  labs(title = "Distribution of Estimated Salary (in euros)", subtitle = "For Customers of a European Bank", x = "Estimated Salary (€)", y = "Count") +
  scale_x_continuous(labels = comma) +
  theme_minimal()
```
(analysis)

**Country of Residence**

```{r geography_summary_visualization, eval = FALSE}
table(bank_churn_clean$Geography)

ggplot(bank_churn_clean, aes(x = Geography)) +
  geom_bar(fill = "steelblue") +
  labs(title = "Distribution of Country of Residence", subtitle = "For Customers of a European Bank", x = "Country of Residence", y = "Count") +
  theme_minimal()
```
(analysis)

**Gender**

```{r gender_summary_visualization, eval = FALSE}
table(bank_churn_clean$Gender)

ggplot(bank_churn_clean, aes(x = Gender)) +
  geom_bar(fill = "#5F9EA0") +
  labs(title = "Number of Customers by Gender", subtitle = "For Customers of a European Bank", x = "Gender", y = "Count") +
  theme_minimal()
```
(analysis)

**Credit Card Ownership**
```{r creditcard_summary_visualization, eval = FALSE}

table(bank_churn_clean$HasCrCard)

ggplot(bank_churn_clean, aes(x = HasCrCard)) +
  geom_bar(fill = "#B0E0E6") +
  labs(title = "Credit Card Ownership", subtitle = "For Customers of a European Bank", x = "Owns a Credit Card", y = "Count") +
  theme_minimal()
```
(analysis)

**Active Bank Membership Status**
```{r activemember_summary_visualization, eval = FALSE}

table(bank_churn_clean$IsActiveMember)

ggplot(bank_churn_clean, aes(x = IsActiveMember)) +
  geom_bar(fill = "#4682B4") +
  labs(title = "Active Bank Membership Status", subtitle = "For Customers of a European Bank", x = "Is Active Member of Bank", y = "Count") +
  theme_minimal()
```
(analysis)

**Customer Churn**
```{r exited_summary_visualization, eval = FALSE}

table(bank_churn_clean$Exited)

ggplot(bank_churn_clean, aes(x = Exited)) +
  geom_bar(fill = "#008080") +
  labs(title = "Customer Churn", subtitle = "For Customers of a European Bank", x = "Exited", y = "Count") +
  theme_minimal()
```
(analysis)

***Step 5: Multivariable Comparisons and Visualizations***

**Churn Status by Credit Score**
```{r creditscore_churn_visualization, eval = FALSE}
ggplot(bank_churn_clean, aes(x = as.numeric(CreditScore), y = factor(Exited))) +
  geom_violin(fill = "#B0E0E6", alpha = 0.6) +
  geom_boxplot(fill = "#5F9EA0") +
  labs(title = "Churn Status by Credit Score", subtitle = "For Customers of a European Bank", x = "Credit Score", y = "Exited") +
  theme_minimal()
```
(analysis)

**Churn Status by Age**
```{r age_churn_visualization, eval = FALSE}
ggplot(bank_churn_clean, aes(x = as.numeric(Age), y = factor(Exited))) +
  geom_violin(fill = "lightblue", alpha = 0.6) +
  geom_boxplot(fill = "steelblue") +
  labs(title = "Churn Status by Age", subtitle = "For Customers of a European Bank", x = "Age", y = "Exited") +
  theme_minimal()
```
**Churn Status by Tenure**
```{r tenure_churn_visualization, eval = FALSE}
ggplot(bank_churn_clean, aes(x = as.numeric(Tenure), y = factor(Exited))) +
  geom_violin(fill = "#B0E0E6", alpha = 0.6) +
  geom_boxplot(fill = "#008080") +
  labs(title = "Churn Status by Tenure", subtitle = "For Customers of a European Bank", x = "Tenure", y = "Exited") +
  theme_minimal()
```
(analysis)

**Churn Status by Number of Products**
```{r numproducts_churn_visualization, eval = FALSE}
ggplot(bank_churn_clean, aes(x = as.numeric(NumOfProducts), y = factor(Exited))) +
  geom_violin(fill = "lightblue", alpha = 0.6) +
  geom_boxplot(fill = "#4682B4") +
  labs(title = "Churn Status by Number of Products", subtitle = "For Customers of a European Bank", x = "Number of Products", y = "Exited") +
  theme_minimal()
```

**Churn Status by Account Balance**
```{r balance_churn_visualization, eval = FALSE}
ggplot(bank_churn_clean, aes(x = as.numeric(Balance), y = factor(Exited))) +
  geom_violin(fill = "lightblue", alpha = 0.6) +
  geom_boxplot(fill = "#4682B4") +
  labs(title = "Churn Status by Balance (in euros)", subtitle = "For Customers of a European Bank", x = "Balance (€)", y = "Exited") +
  theme_minimal()
```
(analysis)

**Churn Status by Estimated Salary**
```{r estsalary_churn_visualization, eval = FALSE}
ggplot(bank_churn_clean, aes(x = as.numeric(EstimatedSalary), y = factor(Exited))) +
  geom_violin(fill = "lightblue", alpha = 0.6) +
  geom_boxplot(fill = "#4682B4") +
  labs(title = "Churn Status by Estimated Salary (in euros)", subtitle = "For Customers of a European Bank", x = "Estimated Salary (€)", y = "Exited") +
  theme_minimal()
```

**Churn Status by Country of Residence**
```{r geography_churn_visualization, eval = FALSE}
library(scales)
ggplot(bank_churn_clean, aes(x = Geography, fill = Exited)) +
  geom_bar(position = "fill") +
  labs(title = "Churn Status by Country of Residence", subtitle = "For Customers of a European Bank", x = "Country", y = "Proportion") +
  scale_y_continuous(labels = percent) +
  theme_minimal()
```

