---
title: "DS210 Project Proposal"
author: "Laina Cleaton"
date: "2025-10-28"
output: html_document
---

**Section 1 - Introduction:**

Customer retention is a critical concern in the banking industry, where competition and customer expectations are both high. Understanding why customers choose to leave a particular bank, or "churn", can provide valuable insights for improving bank services and maintaining customer satisfaction. 

During my internship at Bank of America, I supported the Call Insights function in the Retail Transformation team, focusing on data analytics to enhance the client experience. I enjoyed identifying trends in customer call and complaint data to improve bank service efficiency. This project allows me to build on those skills by applying similar analytical techniques to a different dataset within the banking industry. 

The purpose of this project is to explore the factors that influence customer churn, with the goal of identifying patterns and relationships between customer characteristics and their decision to close their account. The dataset used in this study, titled "Bank Customer Churn," is publicly available from Maven Analytics' Data Playground (https://mavenanalytics.io/data-playground/bank-customer-churn?page=7&pageSize=5), which sourced the data from Kaggle (https://www.kaggle.com/datasets/shubhammeshram579/bank-customer-churn-prediction/data). While the specific bank is not identified, the dataset represents customer information from an unnamed European bank and contains 10,000 observations. 

Each of the 10,000 observations represents a single customer and includes 13 variables capturing demographic, behavioral, and bank account information. Key variables include Geography (the customer's country of residence: France, Germany, or Spain), Tenure (number of years as a customer with the bank), Balance and EstimatedSalary (both in euros, €), and IsActiveMember (whether the customer actively engages with the bank). The outcome variable, Exited, indicates whether or not the customer has closed their account with the bank. 

The data on each customer was collected from the bank's internal records, which means that each row corresponds to one customer's status at the time the data was collected. In this analysis, the variable Exited is the dependent variable representing customer churn, while the remaining variables serve as potential predictors of the decision to churn. 

By examining this dataset, the project aims to uncover trends and key factors that contribute to customer churn. These insights can inform strategies to improve customer retention and guide future efforts in customer relationship management. 

**Section 2 - Exploratory Data Analysis:**

```{r setup_tidyverse, eval=FALSE}
library(tidyverse)
```

```{r read_file, eval = FALSE}
bank_churn_data <- read_csv("Bank_Churn.csv")
View(bank_churn_data)
```

